---
title: "Lab 2"
author: "Adeline Casali"
date: "2024-01-28"
output: word_document
---

Building the Model
```{r}
# Load packages
library(reticulate)
library(tensorflow)
library(keras)

# Preparing data
mtcars <- mtcars
mtcars_x <- mtcars[, c("cyl", "disp", "hp")]
mtcars_x <- array(data = unlist(mtcars_x), 
                  dim = c(32, 3), 
                  dimnames = list(rownames(mtcars_x), colnames(mtcars_x)))
mtcars_y <- mtcars[, "mpg"]

# Building the model
nn_model <- keras_model_sequential() %>%
  layer_dense(units = 1, input_shape = 3, activation = "linear")
nn_model 

# Adding hidden layers
nn_model <- keras_model_sequential() %>% 
  layer_dense(units = 2, input_shape = 3, activation = "relu") %>% 
  layer_dense(units = 1, activation = "linear")
nn_model

# Adding the loss function and optimization
nn_model %>% compile(optimizer = optimizer_adam(learning_rate = 0.2), 
                     loss = "mean_squared_error", 
                     metrics = "mean_absolute_error")

# Training the neural network model
nn_model_training <- nn_model %>% fit(x = mtcars_x, 
                                      y = mtcars_y, 
                                      epochs = 250, 
                                      verbose = FALSE)

# Plotting the process (learning curve)
plot(nn_model_training)
```

1) The predicted miles per gallon of a vehicle with 8 cylinders, a displacement of 250, and horsepower of 200 is 20.67666 mpg. 
```{r}
prediction <- predict(nn_model, array(c(8, 250, 200), dim = c(1, 3)))
prediction
```

2) The loss function specified in the above model is mean squared error. This metric is particularly useful for the loss function in neural networks because it is a smooth and differentiable function, which is essential for optimization algorithms such as gradient descent. 

3) The rank of a tensor refers to the number of axes or dimensions in the tensor. A matrix is a tensor of rank 2 as it has two dimensions (columns and rows). 

4) Vector data is rank 1 tensor data, as it contains one axis. 

5) This tensor is a rank 4 tensor, which is likely that of an image. The shape is (500, 256, 256, 3). In this case, the second dimension would be 256. If this tensor represented a set of color images, there would be 500 images. 













